---
title: "A3: Incarceration"
output: html_document
---

### Introduction

An introduction of the problem domain and a description of the variable(s) you are choosing to analyze (and why!)

Though this analysis I will be examining the rate of incarcerations based on location (rural/urban) in the state of Washington. Additionally, I will examine how race impacts the rate of incarceration in Washington. By limiting the scope of analysis to a singular state, there is more specific insight available into how race can impact incarceration rates. Specifically, I will be examining the rate of Native American incarcerations against the rate of white incarcerations. Due to the relevance of this issue, as the Pacific Northwest and Washington as a whole is a place where many Native Americans reside, the comparison can shed light onto discriminatory behavior. To allow for a more focus scope, I will also only be focusing on the jail population rates. I will be doing this for two reasons: there is little to no information provided in the data frame about prison rates for Native Americans, and the jail rates can provide direct insight into the rate in which Native Americans are arrested. 

### Summary Information

The variables I collected give insight to the rates in which white people are arrested and put in jail compared to the rate Native Americans are arrested. For simplicity, as stated above, I am focusing on specifically Washington state. I used the total jail population per county and state for my analysis as it provides jail numbers rather than rates that were given in the Washington specific table. I created a new data fram containing only WA state data. I found the total Native American jail population over time (1963- present) to be 15551.37 and White jail population to be 222168.9. In order to gauge the percentage of Native Americans and White people incarcerated I then determined the total population of each group. According to the data, in Washington state there are 1,961,627 Native Americans with an incarceration rate of about %0.7927. A total population of 94,711,492 White people reside in Washington with an incarceration rate of about %0.2346. 

```{r}

WA_JAIL_POP <- jail_pop_county_state %>% filter(state == "WA")

native_non_zero_sum <- WA_JAIL_POP %>% 
filter(!is.na(native_jail_pop) & native_pop_15to64 != 0) %>% summarize(sum_native_jail_pop = sum(native_jail_pop))

white_non_zero_sum <- WA_JAIL_POP %>% 
filter(!is.na(white_jail_pop) & white_jail_pop != 0) %>% summarize(sum_white_jail_pop = sum(white_jail_pop))

wa_white_pop <- WA_JAIL_POP %>% filter(!is.na(white_pop_15to64) & white_pop_15to64 != 0) %>% summarise(sum_white_pop = sum(white_pop_15to64))

wa_native_pop <- WA_JAIL_POP %>% filter(!is.na(white_pop_15to64) & white_pop_15to64 != 0) %>% summarise(sum_white_pop = sum(white_pop_15to64))

native_percentage <- native_non_zero_sum / wa_native_pop

white_percentage <- white_non_zero_sum / wa_white_pop

wa_urbanicity_summary <- WA_JAIL_POP %>%
group_by(urbanicity) %>%
summarise(total_jail_population = sum(total_jail_pop, na.rm = TRUE))

incarceration_rate_small_mid <- max(wa_urbanicity_summary$total_jail_population) 

incarceration_rate_urbanicity <- wa_urbanicity_summary %>%
filter(total_jail_population == max(total_jail_population)) %>%
pull(urbanicity)

SMALL_MID_DATA <- WA_JAIL_POP %>% filter(urbanicity == "small/mid")

native_small_mid <- SMALL_MID_DATA %>% filter(!is.na(native_pop_15to64) & native_pop_15to64 != 0) %>% summarise(native_pop_small_mid = sum(native_pop_15to64))

white_small_mid <- SMALL_MID_DATA %>% filter(!is.na(white_pop_15to64) & white_pop_15to64 != 0) %>% summarise(white_pop_small_mid = sum(white_pop_15to64))

total_small_mid <- SMALL_MID_DATA %>% filter(!is.na(total_pop_15to64) & total_pop_15to64 != 0) %>% summarise(total_pop_small_mid = sum(total_pop_15to64))

native_small_mid_percentage <- native_small_mid / total_small_mid

white_small_mid_percentage <- white_small_mid / total_small_mid
```

WA_JAIL_POP: the population of washington incarceration rather than the rate of incarceration

### The Dataset

Who collected the data?  
How was the data collected or generated?  
Why was the data collected?  
How many observations (rows) are in your data?  
How many features (columns) are in the data?  
What, if any, ethical questions or questions of power do you need to consider when working with this data?  
What are possible limitations or problems with this data?   (at least 200 words)

This data was collected by Vera in order to provide insight into the rates on incarceration as well as details into how different racial groups are disproportionately incarcerated. Through the sources 

### Trends Over Time Chart

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The first chart that you will create and include will show the trend over time of your variable/topic. Think carefully about what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- Show more than one, but fewer than ~10 trends
  - This may mean showing the same measure for different locations or different racial groups. Think carefully about a meaningful comparison of locations (e.g., the top 10 counties in a state, top 10 states, etc.)
- You must have clear x and y axis labels
- The chart needs a clear title 
- You need a legend for your different line colors and a clear legend title

When we say "clear" or "human readable" titles and labels, that means that you should not just display the variable name.

Here's an example of how to run an R script inside an RMarkdown file:

``````{r, echo = FALSE, code = readLines("chart1_example.R")}
```

### Variable Comparison Chart

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The second chart that you will create and include will show how two different (continuous) variables are related to one another. Again, think carefully about what such a comparison means and what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- You must have clear x and y axis labels
- The chart needs a clear title 
- If you choose to add a color encoding (not required), you need a legend for your different color and a clear legend title

### Map

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The last chart that you will create and include will show how a variable is distributed geographically. Again, think carefully about what such a comparison means and what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- Your map needs a title
- Your color scale needs a legend with a clear label
- Use a map based coordinate system to set the aspect ratio of your map

